\documentclass[math, info]{cours}
\title{Homework 1\\{\small Convex Optimization}}
\author{Matthieu Boyer}

\begin{document}
\maketitle

\section{Exercises}
\subsection*{Convex Sets}
\begin{itemize}
	\item Let $\mC = \left\{ x \in \R_{+}^{2} \mid x_{1}x_{2} \geq 1 \right\}$.
		We can see that $\mC = \left\{ (x_{1}, x_{2}) \in \R_{+*}^{2} \mid x_{1} \geq \frac{1}{x_{2}} \right\}$.
		Thus, $\mC$ is the epigraph of $x \mapsto \frac{1}{x}$ which is convex since its derivative is $x \mapsto \frac{1}{x^{4}}$.
		\item In general, this set is not convex. Indeed, for sets in $\R$, take $S = \{-1, 1\}, T = \{0\}$.
		Then the set of points closer to $S$ than to $T$ is $\{x \in \R \mid x\leq -0.5 \lor x \geq 0.5\}$.
			\item Let us take $x_{1}, x_{2} \in \left\{ x \mid x + S_{2} \subseteq S_{1} \right\}$.
		For $x = \lambda x_{1} + \left( 1 - \lambda \right)x_{2}$ consider $x + y$ for $y \in S_{2}$:
		\begin{equation*}
			x + y = \lambda x_{1} + \left(1 - \lambda \right)x_{2} + y = \lambda(x_{1} + y) + (1 - \lambda x_{2} + y)
		\end{equation*}
		Since $x_{i} + S_{2}\subseteq S_{1}$ for $i = 1, 2$, and since $S_{1}$ is convex, for all $y \in S_{2}$ the above sum is in $S_{1}$ and thus $x + S_{2} \subseteq S_{2}$, and our set is convex.
	\item Let $x_{1}, x_{2} \in \left\{ x \mid \exists y \in S_{2}, x + y \in S_{1} \right\}$.
		Then let $y_{1}, y_{2}$ be associated points to $x_{1}, x_{2}$.
		For $x = \lambda x_{1} + \left( 1 - \lambda \right)x_{2}$ and $y = \lambda y_{1} + (1 - \lambda) y_{2}$.
		Then, since $S_{2}$ is convex, $y \in S_{2}$.
		Moreover:
		\begin{equation*}
			x + y = \underbrace{\lambda \underbrace{(x_{1} + y_{1})}_{\in S_{1}} + \left( 1 - \lambda \right)\underbrace{(x_{2} + y_{2})}_{\in S_{1}}}_{\in S_{1} \text{ since } S_{1} \text{ is convex}}
		\end{equation*}
		Then, our set is convex.
\end{itemize}

\subsection*{Convex Functions}
\begin{itemize}
	\item The hessian of $f$ at $(x,y)$ is the matrix $\begin{pmatrix}
			0 & 1\\
			1 & 0
		\end{pmatrix}$.
		On $\R^{2}$ this matrix is not positive nor negative and thus the function is neither convex, or concave.
		However, the upper level sets are convex (from our first example) and thus the function is quasi-concave.
	\item On $\R_{+*}^{2}$, the hessian of the function is positive semidefinite and thus the function is convex.
	\item The hessian matrix of $f$ is the matrix$\begin{pmatrix}
			0 & -\frac{1}{x_{2}^{2}}\\
			-\frac{1}{x_{2}^{2}} & \frac{2x_{1}}{x_{2}^{3}}
		\end{pmatrix}$
		Since its determinant is $<0$ the Hessian is not positive semi-definite and the function is not convex.
		However, its sublevel sets are defined by the equations $x_{1} \leq \alpha x_{2}$ and are thus convex (since half-planes).
		Thus, $f$ is quasi-convex.
	\item Let us define the Löwner order $\preceq$ as the partial order defined by the convex cone of positive semi-definite matrices.
		We know that:
		\begin{equation*}
			A \preceq B \Rightarrow B^{-1} \preceq A^{-1} \forall A, B \in S_{++}^{n}
		\end{equation*}
		From this, we know that for all $t \leq 1$:
		\begin{equation*}
			\left( \left( 1 - t \right) X + tY \right)^{-1} \preceq \left( 1 - t \right)X^{-1} + tY^{-1}
		\end{equation*}
		By linearity of the trace, we can now see that $X \mapsto \Tr(X^{-1})$ is convex.
\end{itemize}

\subsection*{Fenchel Conjugate}
\begin{itemize}
	\item We have:
		\begin{equation*}
			f^{*}(y) = \sup_{x}\left( \transpose{x}y -\norm{x} \right) = \begin{cases}
				0 & \text{if } \norm{y}_{*} \leq 1\\
				\infty &  \text{otherwise}
			\end{cases}
		\end{equation*}
		To show this, let $y$ such that $\norm{y}_{*} = \sup_{\norm{x} \leq 1} \norm{\transpose{x}y} \leq 1$.
		Then by Cauchy-Schwarz inequality, we know that $\transpose{x}y \leq \norm{x}\norm{y}_{*}\leq \norm{x}$ and the term in the supremum is always $\leq 0$.
		If $\norm{y}_{*} > 1$ however, then there exists $z$ such that $\norm{z} \leq 1$ such that $\transpose{z}y > 1$.
		Taking $x = tz$ in the supremum, we know that $f^{*}(y) \geq t\left( \transpose{z}y -\norm{z} \right)$ which goes to infinity with $t \to \infty$.
		Thus, the fenchel conjugate of a norm is the convex indicator of the unit ball of the dual norm.
	\item Let us denote the infimal convolution of $g, h$ by $g \square h$.
		Then we will show that:
		\begin{equation*}
			\left( g \square h \right)^{*} = \left( g^{*} + h^{*} \right)
		\end{equation*}
		Note that:
		\begin{equation*}
			\begin{aligned}
				\left( g\square h \right)^{*}(\alpha) =& \sup_{x, y} \left\{ \transpose{x}\alpha - g(y) - h(x - y) \right\}\\
			=& \sup_{x_{1}, x_{2}} \left\{ \transpose{\left( x_{1} + x_{2} \right)}\alpha - g(x_{1}) - h(x_{2}) \right\}\\
			=& \sup_{x_{1}, x_{2}} \left\{ \transpose{x_{1}}\alpha - g(x_{1})\right\} + \sup_{x_{2}}\left\{ \transpose{x_{2}}\alpha - h(x_{2}) \right\}\\
			=& g^{*} + h^{*}
		\end{aligned}
		\end{equation*}
		Given $g = \norm{\cdot}_{1}$ and $h = \frac{1}{2\alpha}\norm{\cdot}_{2}^{2}$.
		Let $x = u + v$. Then:
		\begin{equation*}
			f(x) = \inf_{v} \left\{ g(x - v) + h(v) \right\}
		\end{equation*}
		Substituting the expressions, we get:
		\begin{equation*}
			\begin{aligned}
				f(x) =& \inf_{v}\left\{ \norm{x - v}_{1} + \frac{1}{2\alpha}\norm{v}_{2}^{2} \right\}\\
				=& \sum_{i = 1}^{n} \inf_{v_{i}} \left\{ \abs{x_{i} - v_{i}} + \frac{1}{2\alpha}v_{i}^{2} \right\}
			\end{aligned}
		\end{equation*}
		We will now compute the infima independently. We have two cases:
		\begin{enumerate}
			\item $v_{i} \leq x_{i}$: Let $\phi(v_{i}) = x_{i} - v_{i} + \frac{1}{2\alpha}v_{i}^{2}$. We want to find a minimum for $\phi$, which is found at $v_{i} = \alpha$. This solution is valid if $\alpha \leq x_{i}$.
			\item $v_{i} > x_{i}$: Let $\phi(v_{i} = \frac{1}{2\alpha} + v_{i} - x_{i}$. We want to find a minimum for $\phi$, which is at $v_{i} = -\alpha$. This solution is valid if $\alpha > x_{i}$.
		\end{enumerate}
		Then, we have three possible cases for the optimal $v_{i}$:
		\begin{enumerate}
			\item If $x_{i} \geq \alpha$, $v_{i} = \alpha$
			\item If $x_{i} \leq -\alpha$, $v_{i} = -\alpha$
			\item If $-\alpha < x_{i} < \alpha$, $v_{i} = x_{i}$.
		\end{enumerate}
		Plugging this into $f$:
		\begin{equation*}
			\boxed{%
			f(x) = \sum_{i = 1}^{n} \left( \begin{cases}
					\frac{1}{2\alpha}x_{i}^{2} & \abs{x_{i}}\leq \alpha\\
					\abs{x_{i} - \alpha} - \frac{\alpha}{2} & \abs{x_{i}} > \alpha
		\end{cases}\right)}
		\end{equation*}
	\item We want to compute:
		\begin{equation*}
			\ell^{*}(y) = \sup_{z\in \R}\left\{ yz - \ell(z) \right\} \text{ where } \ell: z\mapsto \log\left( 1 + e^{-z} \right)
		\end{equation*}
		We differentiate $\phi(z)$ the function in the supremum:
		\begin{equation*}
			\phi'(z) = y - \frac{1}{1 + e^{z}}
		\end{equation*}
		Then, we find $z = \log\left( \frac{1 - y}{y} \right)$, which only makes sense for $y \in ]0, 1[$.
		Finally, we compute:
		\begin{equation*}
			\begin{aligned}
				\ell^{*}(y) =& y\log\left( \frac{1-y}{y} \right) + \log(1- y)\\
				=& y\log(1 - y) - y\log(y) + \log(1 - y)\\
				=& \log(1 - y) - y\log(y)
			\end{aligned}
		\end{equation*}
		In the end,
		\begin{equation*}
			l^{*}(y)
		\end{equation*}<++>
\end{itemize}

\end{document}
